---
title: "STAA57 - WorkSheet 3"
date: "  "
author: "Name:    , ID#:   "
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Goal**: Master tidy data organization and subsetting. 

1. Reshape the following data so that they are *tidy*

```{r}
my_df = data_frame( group = c("A","B","C"), 
                    drug_X = c(130, 135, 140),
                    drug_Y = c(NA, 150, 135))
```


For the remaining questions use the *dinesafe* data. 
```{r, include = FALSE}
library(tidyverse)
dinesafe = read_csv("data/dinesafe.csv") 
```

2. Arrange infractions from largest to smallest amount fined. Return the inspection ID, establishment name, amount fined, and infraction details.

3. Find the *restaurant* infractions with the 10 highest fines. Return the inspection ID, establishment name, amount fined, and infraction details.


4. Find how many (unique) inspections took place in 2017.

5. Looking at distinct values of the variable SEVERITY, there are two types of values labeled <NA> and "NA - Not Applicable". Can you explain the difference between the two? 

```{r}
dinesafe %>% distinct(SEVERITY)
```

6. Split SEVERITY into two variables: a severity code (SEV_CODE) and a severity description (SEV_DESC): i.e. `separate()` the values of SEVERITY along the hyphen (--).

7. Find one of my favourite spots in TO: it was inspected on 2018-03-20, got a Pass, and has the 4th smallest ID \# out of all the other establishments inspected on that day. Write R code that outputs the name of the place.  
(Hint: Use `top_n()` with *negative* row index.)


8. Create a list of all *unique* Tim Hortons establishments in Toronto. The name can be either "TIM HORTONS" or "Tim Hortons", and you should use `select()` with `starts_with()`  to return all variables starting with "ESTABLISHMENT".






